---
title: "Adding Bio Tooltips to R Markdown reports"
output:
  rmarkdown::html_vignette:
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Adding Bio Tooltips to R Markdown reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(biotooltips)
use_bio_tooltips(
  modules = c("gene", "chemical"),
  include_optional_visual_deps = FALSE
)
```

`biotooltips` adds small HTML spans to R Markdown output and lets the browser-side Bio Tooltips JavaScript library do the lookup and rendering.

## Inline prose

Genes can be included inline:

```{r inline-demo, results='asis'}
cat("The tumour suppressor ", gene_tt("TP53", species = "human"), " is central to DNA damage responses.", sep = "")
```

Chemicals can use stable identifiers:

```{r chemical-demo, results='asis'}
cat("A chemical example: ", chem_tt("aspirin", query = "2244", scope = "pubchem"), ".", sep = "")
```

## Tables

```{r table-demo, results='asis'}
top_genes <- data.frame(
  symbol = c("TP53", "BRCA1", "GADD45A"),
  log2FoldChange = c(2.1, -1.4, 1.2),
  padj = c(0.0004, 0.002, 0.01)
)

top_genes |>
  gene_column(symbol, species = "human") |>
  bt_kable(include_setup = FALSE)
```

The important detail is that HTML escaping must be disabled for the table cell containing tooltip spans. `bt_kable()` does this by default.

## Assets

`use_bio_tooltips()` uses vendored `bio-tooltips` 1.0.0 browser assets by
default, so building this vignette does not download JavaScript or CSS files.
To use jsDelivr-hosted assets in a rendered report, opt in explicitly:

```{r cdn-demo, eval=FALSE}
use_bio_tooltips(cdn = TRUE, version = "1.0.0")
```

## Volcano plot

Interactive plotting libraries usually own their hover labels, so the most
reliable pattern is to let Plotly handle the plot hover event and update a real
HTML Bio Tooltip span next to the plot.

```{r volcano-demo, message=FALSE, warning=FALSE}
if (!requireNamespace("plotly", quietly = TRUE) ||
    !requireNamespace("htmlwidgets", quietly = TRUE)) {
  htmltools::tags$p(
    "Install the optional plotly and htmlwidgets packages to render this interactive example."
  )
} else {
set.seed(42)

volcano_genes <- c(
  "TP53", "BRCA1", "BRCA2", "EGFR", "MYC", "PTEN", "KRAS", "NRAS",
  "BRAF", "CDK1", "CCND1", "CDKN1A", "CDKN2A", "MDM2", "RB1", "ATM",
  "ATR", "CHEK1", "CHEK2", "GADD45A", "VEGFA", "HIF1A", "IL6", "TNF",
  "CXCL8", "STAT3", "JAK2", "AKT1", "MTOR", "PIK3CA", "FOXO3", "ESR1",
  "AR", "ERBB2", "MET", "ALK", "ROS1", "NTRK1", "RET", "APC",
  "CTNNB1", "SMAD4", "TGFB1", "MMP2", "MMP9", "COL1A1", "COL3A1",
  "FN1", "VIM", "CDH1", "EPCAM", "SOX2", "NANOG", "POU5F1", "MKI67",
  "AURKA", "TOP2A", "BCL2", "BAX"
)

volcano <- data.frame(
  symbol = volcano_genes,
  log2FoldChange = rnorm(length(volcano_genes), sd = 0.75),
  padj = runif(length(volcano_genes), min = 0.08, max = 0.95)
)

interesting <- data.frame(
  symbol = c("MYC", "EGFR", "VEGFA", "IL6", "MMP9", "MKI67",
             "TP53", "BRCA1", "CDKN1A", "PTEN", "CDH1", "BAX"),
  log2FoldChange = c(2.6, 1.9, 1.7, 2.2, 1.5, 2.9,
                     -2.1, -1.8, -1.4, -1.7, -2.4, -1.3),
  padj = c(0.00012, 0.0018, 0.006, 0.0009, 0.014, 0.00004,
           0.0007, 0.003, 0.018, 0.009, 0.0002, 0.022)
)

volcano[match(interesting$symbol, volcano$symbol), c("log2FoldChange", "padj")] <-
  interesting[c("log2FoldChange", "padj")]

volcano$neg_log10_padj <- -log10(volcano$padj)
volcano$direction <- ifelse(
  volcano$padj < 0.05 & volcano$log2FoldChange >= 1,
  "Up-regulated",
  ifelse(
    volcano$padj < 0.05 & volcano$log2FoldChange <= -1,
    "Down-regulated",
    "Not significant"
  )
)
volcano$direction <- factor(
  volcano$direction,
  levels = c("Up-regulated", "Down-regulated", "Not significant")
)
volcano$label <- ifelse(volcano$symbol %in% interesting$symbol, volcano$symbol, "")

volcano_plot <- plotly::plot_ly(
  volcano,
  x = ~log2FoldChange,
  y = ~neg_log10_padj,
  key = ~symbol,
  type = "scatter",
  mode = "markers+text",
  text = ~label,
  textposition = "top center",
  color = ~direction,
  colors = c(
    "Up-regulated" = "#b91c1c",
    "Down-regulated" = "#2563eb",
    "Not significant" = "#7a7f87"
  ),
  marker = list(size = 9, opacity = 0.82, line = list(width = 0)),
  textfont = list(size = 10, color = "#24292f")
) |>
  plotly::layout(
    xaxis = list(title = "log2 fold change", zeroline = TRUE),
    yaxis = list(title = "-log10 adjusted p-value"),
    legend = list(orientation = "h", x = 0, y = 1.12),
    margin = list(t = 60),
    shapes = list(
      list(type = "line", x0 = -1, x1 = -1, y0 = 0, y1 = 4.6,
           line = list(color = "#b7bdc5", width = 1, dash = "dot")),
      list(type = "line", x0 = 1, x1 = 1, y0 = 0, y1 = 4.6,
           line = list(color = "#b7bdc5", width = 1, dash = "dot")),
      list(type = "line", x0 = -3.2, x1 = 3.2, y0 = -log10(0.05), y1 = -log10(0.05),
           line = list(color = "#b7bdc5", width = 1, dash = "dot"))
    )
  )

bt_plotly_gene_hover(volcano_plot, species = "human", include_setup = FALSE)
}
```

Plotly still owns the hover event, but its default hover card is suppressed.
`bt_plotly_gene_hover()` reads the gene symbol from Plotly's `key` field and
updates a tiny fixed-position Bio Tooltip span at the cursor.

## Experimental auto-linking

If prose is already rendered, `auto_gene_tooltips()` can wrap a supplied vocabulary of gene symbols in a constrained selector. This should be used carefully because some gene symbols are ordinary words.

```{r auto-demo, eval=FALSE}
auto_gene_tooltips(
  genes = c("TP53", "BRCA1", "GADD45A"),
  species = "human",
  selector = ".results-section"
)
```
