---
title: "Adding Bio Tooltips to R Markdown reports"
output:
  rmarkdown::html_vignette:
    self_contained: false
vignette: >
  %\VignetteIndexEntry{Adding Bio Tooltips to R Markdown reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(biotooltips)
use_bio_tooltips(modules = c("gene", "chemical"))
```

`biotooltips` adds small HTML spans to R Markdown output and lets the browser-side Bio Tooltips JavaScript library do the lookup and rendering.

## Inline prose

Genes can be included inline:

```{r inline-demo, results='asis'}
cat("The tumour suppressor ", gene_tt("TP53", species = "human"), " is central to DNA damage responses.", sep = "")
```

Chemicals can use stable identifiers:

```{r chemical-demo, results='asis'}
cat("A chemical example: ", chem_tt("aspirin", query = "2244", scope = "pubchem"), ".", sep = "")
```

## Tables

```{r table-demo, results='asis'}
top_genes <- data.frame(
  symbol = c("TP53", "BRCA1", "GADD45A"),
  log2FoldChange = c(2.1, -1.4, 1.2),
  padj = c(0.0004, 0.002, 0.01)
)

top_genes |>
  gene_column(symbol, species = "human") |>
  bt_kable(include_setup = FALSE)
```

The important detail is that HTML escaping must be disabled for the table cell containing tooltip spans. `bt_kable()` does this by default.

## Experimental auto-linking

If prose is already rendered, `auto_gene_tooltips()` can wrap a supplied vocabulary of gene symbols in a constrained selector. This should be used carefully because some gene symbols are ordinary words.

```{r auto-demo, eval=FALSE}
auto_gene_tooltips(
  genes = c("TP53", "BRCA1", "GADD45A"),
  species = "human",
  selector = ".results-section"
)
```
